Constraints on Cosmic Rays, Magnetic Fields, and Dark Matter from Gamma-Ray Observations of the Coma Cluster of Galaxies with VERITAS and \Fermi

### 2011-12-14 ms.tex
F_min discussion clarified

We start by adopting a simplified analytical model that assumes a power-law CR spectrum and a
constant relative CR pressure, i.e., we adopt the {\em isobaric model of CRs} following the approach
of \citet{article:PfrommerEnsslin:2004b}. To be independent of additional assumptions and in line
with earlier work in the literature, we do not impose a low-momentum cutoff, $q$, on the CR
distribution function, i.e., we adopt $q=0$. Since {\em a priori}, the CR spectral index is
unconstrained\footnote{The hadronic interaction physics guarantees that the CR spectral index
coincides with that of the resulting pion-decay gamma-ray emission.}, we vary it in the range
$2.1<\alpha<2.5$, which is compatible with the radio spectral index of the giant radio halo of the
Coma cluster after accounting for the spectral steepening at frequencies $\nu\sim5~\rmn{GHz}$ due to
the Sunyaev-Zel'dovich effect \citep{article:Ensslin:2002,
article:PfrommerEnsslin:2004b}\footnote{Assuming a magnetic field of 1 $\mu$G, the CR protons
responsible for the GHz radio emitting electrons have an energy of $\sim100$ GeV and are $\sim$ 20
times less energetic than those CR protons responsible for 200-GeV gamma-ray emission.}. To model
the thermal pressure, we adopt the electron density profile for the Coma cluster that has been
inferred from ROSAT X-ray observations \citep{article:BrielHenryBohringer:1992} and use a constant
temperature of $kT= 8.25$~keV throughout the virial region.

@@

The results for the minimum gamma-ray flux $F_{\gamma,\rmn{min}}(>220~\rmn{GeV})$ and the minimum
relative CR pressure, $X_{\CR,\,\rmn{min}} = X_\CR F_{\gamma,\rmn{min}}/F_{\gamma,\rmn{iso}}$ are
shown in Table \ref{table:constraints}, where $F_{\gamma,\rmn{iso}}$ is the gamma-ray flux in the
simplified model introduced in \S\ref{sec:simple}. Even in the most constraining cases, these are a
factor of about 60 below the VERITAS upper limits (adopting $\alpha=2.1$, $<0.2^{\circ}$) and a
factor of about 20 below the \Fermi upper limits (adopting $\alpha=2.3$, $<0.4^{\circ}$). Note
that these minimum gamma-ray fluxes are sensitive to a running of the CR proton spectral index with
energy as a result of, e.g., momentum-dependent diffusion: assuming a plausible value for the
central magnetic field of Coma of 5~$\mu$G \citep{article:Bonafede_etal:2010}, the radio halo
emission at GHz frequencies is dominated by the electrons with an energy $E_\rmn{e} \sim 2.5$ GeV
(which corresponds to proton energies $E_\rmn{p} \sim 40$ GeV). Gamma rays with an energy of $200$
GeV are produced by CR protons with an energy of $E_\rmn{p} \sim 1.6$ TeV -- a factor of 50 higher
than those probed by radio halo observations. A steepening of the CR proton spectral index of 0.2
between 40 GeV and 1.6 TeV would imply a decrease in the minimum gamma-ray flux by a factor of two.

>>>>

We start by adopting a simplified analytical model that assumes a power-law CR spectrum and a
constant relative CR pressure, i.e., we adopt the {\em isobaric model of CRs} following the approach
of \citet{article:PfrommerEnsslin:2004b}. To be independent of additional assumptions and in line
with earlier work in the literature, we do not impose a low-momentum cutoff, $q$, on the CR
distribution function, i.e., we adopt $q=0$. Since {\em a priori}, the CR spectral index is
unconstrained\footnote{The hadronic interaction physics guarantees that the CR spectral index
  coincides with that of the resulting pion-decay gamma-ray emission \citep[see discussion
  in][]{article:PfrommerEnsslin:2004b}.}, we vary it in the range $2.1<\alpha<2.5$, which is
compatible with the radio spectral index of the giant radio halo of the Coma cluster after
accounting for the spectral steepening at frequencies $\nu\sim5~\rmn{GHz}$ due to the
Sunyaev-Zel'dovich effect \citep{article:Ensslin:2002,
  article:PfrommerEnsslin:2004b}\footnote{Assuming a magnetic field of 1 $\mu$G, the CR protons
  responsible for the GHz radio emitting electrons have an energy of $\sim100$ GeV and are $\sim$ 20
  times less energetic than those CR protons responsible for 200-GeV gamma-ray emission.}. To model
the thermal pressure, we adopt the electron density profile for the Coma cluster that has been
inferred from ROSAT X-ray observations \citep{article:BrielHenryBohringer:1992} and use a constant
temperature of $kT= 8.25$~keV throughout the virial region.

@@

The results for the minimum gamma-ray flux $F_{\gamma,\rmn{min}}(>220~\rmn{GeV})$ and the minimum
relative CR pressure, $X_{\CR,\,\rmn{min}} = X_\CR F_{\gamma,\rmn{min}}/F_{\gamma,\rmn{iso}}$ are
shown in Table \ref{table:constraints}, where $F_{\gamma,\rmn{iso}}$ is the gamma-ray flux in the
simplified model introduced in \S\ref{sec:simple}. \C{Even in the most constraining cases and
  assuming $\alpha\leq 2.3$, these are a factor of $\sim 60$ below the VERITAS upper limits (for
  $\alpha=2.1$, $<0.2^{\circ}$) and a factor of $\sim 20$ below the \Fermi upper limits (for
  $\alpha=2.3$, $<0.4^{\circ}$)}. Note that these minimum gamma-ray fluxes are sensitive to a
running of the CR proton spectral index with energy as a result of, e.g., momentum-dependent
diffusion: assuming a plausible value for the central magnetic field of Coma of 5~$\mu$G
\citep{article:Bonafede_etal:2010}, the radio halo emission at GHz frequencies is dominated by the
electrons with an energy $E_\rmn{e} \sim 2.5$ GeV (which corresponds to proton energies $E_\rmn{p}
\sim 40$ GeV). Gamma rays with an energy of $200$ GeV are produced by CR protons with an energy of
$E_\rmn{p} \sim 1.6$ TeV -- a factor of 50 higher than those probed by radio halo observations. A
steepening of the CR proton spectral index of 0.2 between 40 GeV and 1.6 TeV would imply a decrease
in the minimum gamma-ray flux by a factor of two.
